Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-17 paperDistance/paper.tex
math tweaks

% text macros
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\panstarrs}{\project{Pan\acronym{STARRS}}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}
\newcommand{\given}{\,|\,}
% math macros
\DeclareMathOperator*{\argmax}{arg\,max}

@@

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{true}, \sigma^2_{\varpi}) \, p(\varpi_{true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (CITE TRI--CORYN), or relying on stellar
models, we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star, a prior pdf for the distance
to that star.

@@

\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \;\;\;\; &\vec{Y} = [10^{0.2\,M_{J,n} + 2}, C_n] \\
\mathrm{data \, vector} \;\;\;\; &\vec{y} = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \;\;\;\; &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \;\;\;\; &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{assumptions} \;\;\;\; &I = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}

@@

To infer $\varpi_{true}$ from the likelihood of the \gaia\ data, Equation \ref{eq:gaialike}, we first do the inference in the 2D space of the \cmd

@@

\begin{eqnarray}
p(\vec{Y}_0 \given \vec{y}, \alpha) &\rightarrow& p(\varpi_{true} \given y, \alpha) \, f(\varpi_{true}) \\
f(\varpi_{true}) &=& \begin{cases}
              1: -2 < \mathrm{log}_{10} \varpi_{true} < 2\\
              0: \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}

>>>>

% text macros
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\panstarrs}{\project{Pan\acronym{STARRS}}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

% math macros
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\given}{\,|\,}
\newcommand{\true}{\mathrm{true}}

@@

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{\true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (CITE TRI--CORYN), or relying on stellar
models, we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star, a prior pdf for the distance
to that star.

@@

\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\vec{Y} = [10^{0.2\,M_{J,n} + 2}, C_n] \\
\mathrm{data \, vector} \quad &\vec{y} = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{assumptions} \quad &I = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}

@@

To infer $\varpi_{\true}$ from the likelihood of the \gaia\ data, Equation \ref{eq:gaialike}, we first do the inference in the 2D space of the \cmd

@@

\begin{eqnarray}
p(\vec{Y}_0 \given \vec{y}, \alpha) &\rightarrow& p(\varpi_{\true} \given y, \alpha) \, f(\varpi_{\true}) \\
f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
